---
title: "Using Rcompadre with the tidyverse"
author: "Patrick Barks & Owen Jones"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Rcompadre with the tidyverse}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setupDarwin, include=FALSE, eval = Sys.info()[["sysname"]] == "Darwin"}
# The following line seems to be required by pkgdown::build_site() on my machine,
# but causes build to break with R-CMD-CHECK on GH
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

## Introduction

Rcompadre includes methods for a variety of functions in the "[Tidyverse](https://www.tidyverse.org/)", a popular group of R packages geared toward data analysis, including [dplyr](https://dplyr.tidyverse.org/) (a grammar of data manipulation) and [ggplot2](https://ggplot2.tidyverse.org/) (a grammar of graphics). This vignette covers manipulation of `CompadreDB` objects using `dplyr`, and a few examples of plotting `CompadreDB` objects using `ggplot2.` Users new to these packages may wish to check out more introductory materials (e.g. the [dplyr vignette](https://CRAN.R-project.org/package=dplyr), or the [Data Visualization chapter](https://r4ds.had.co.nz/data-visualisation.html) of _R for Data Science_).


## Preliminaries

We'll begin by loading a few packages.

```{r message=FALSE}
library(Rcompadre)
library(dplyr)
library(ggplot2)
library(popdemo)
```

## Introduction to piping

We can take advantage of piping. The pipe (`%>%`) (from the [magrittr](https://magrittr.tidyverse.org/) package), passes an object on the left to the first argument of a function on the right. There is also a native pipe (after R v. 4.1.0) but the `magrittr` pipe has some advantages because it has a placeholder element for when one does not want the left hand side result to go to the *first* argument of the right hand expression.

```{r results=FALSE}
y <- c(0.2, 4.1, 3.7)

mean(y) # 'normal' expression
y %>% mean() # piped expression
```

Though we generally don't need to, we can explicitly refer to the object on the left using a dot (`"."`).

```{r results=FALSE}
y %>% mean() # dot is implicit
y %>% mean(x = .) # dot is explicit
```

The dot notation is particularly helpful if we want to pass the object on the left to an argument _other_ than the first one.

```{r results=FALSE}
x <- 1:3
y %>% data.frame(col1 = x, col2 = .) # use dot to pass object to second argument
```

Using a pipe in a single line isn't that helpful. The real benefits come when we use pipes in multi-line expressions to carry out a sequence of related operations.

#### Piping with Rcompadre

Let's say we want to remove all rows where matA contains missing values (`NA`). We can use `cdb_flag()` to add a column checking for `NA` (column "check_NA_A"), and then use `filter()` to remove those rows. Here are two approaches without piping:

```{r eval=FALSE}
# approach 1 (nested functions)
compadre_use <- filter(cdb_flag(Compadre), check_NA_A == FALSE)

# approach 2 (intermediate step)
compadre_flag <- cdb_flag(Compadre)
compadre_use <- filter(compadre_flag, check_NA_A == FALSE)
```

and here's the equivalent piped sequence:

```{r}
compadre_use <- Compadre %>%
  cdb_flag() %>% # first argument is Compadre, from previous line
  subset(check_NA_A == FALSE) # first argument is output of cdb_flag()
```

The advantage of piping here is that we don't have to use nested functions (`filter(cdb_flag())`), and we don't have to create object names for every intermediate step in our analysis. An additional alternative is to use `subset()` instead of `filter()`.


## The `mutate` function

The `mutate()` function in dplyr adds one or more new columns to a data frame (or in our case, a `CompadreDB` object). Often the new column will be based on some transformation of one or more of the existing columns.

Let's say we want to conduct an analysis comparing Nordic countries to the rest of Europe. We can use `filter()` to limit the database to Europe, and then use `mutate()` to create a new column identifying rows from Nordic countries.

```{r, warning=FALSE}
compadre_euro <- Compadre %>%
  filter(Continent == "Europe") %>%
  mutate(Nordic = Country %in% c("NOR", "SWE", "DNK", "ISL", "FIN"))
```


#### Using `mutate` with Rcompadre functions that return vectors

A variety of `Rcompadre` functions take a `CompadreDB` object as their first argument and return a vector. To use these functions within `mutate()` in a piped sequence, we generally need to explicitly refer to the object on the left side of the pipe with a dot (`"."`).

Here's an example with the Rcompadre functions `mpm_has_active()` and `cdb_id_studies()`, each of which take a `CompadreDB` object and return a vector.

```{r}
compadre_use <- Compadre %>%
  mutate(has_active = mpm_has_active(.)) %>%
  filter(has_active == TRUE) %>%
  mutate(StudyID = cdb_id_studies(.))
```

In the example above, the pipe passes a `CompadreDB` object to the first arguments of `mutate()` and `subset()`, respectively. The dot for the first argument is implicit. Then, within `mutate()`, we use the dot again (explicitly this time) to pass the `CompadreDB` object to `mpm_has_active()` or `cdb_id_studies()`.

We can also use the dot notation approach to extract components from the `CompadreMat` objects in the column `mat`, such as `matA`, `matU`, `matF`, `matC`, `matrixClass`, `MatrixClassOrganized`, or `MatrixClassAuthor.` Each of these components can be accessed using an accessor function of the same name. Here's how to extract a list of `matU` and a list of `MatrixClassOrganized` from each row of the database, and add these lists to a `CompadreDB` object as new columns:

```{r}
compadre_unnest <- Compadre %>%
  mutate(
    mat_U = matU(.),
    m_class_organized = MatrixClassOrganized(.)
  )
```

#### Vectorizing within `mutate` with the apply functions

Just like creating a new column with `$` (e.g. `compadre$new_col <- ...`), each expression within `mutate()` must return a vector of the same length as the number of rows in the data frame, or return a single value which will be recycled for all rows. For some operations this requires vectorization.

Let's say we want to calculate the population growth rate for every matrix in the database. We can use the `eigs()` function in the [popdemo](https://CRAN.R-project.org/package=popdemo) package, but it can only take a single matrix at a time.

To apply `eigs()` to a list of matrices we can use `sapply()`. But first we need to remove matrices with missing values, because these will cause `eigs()` to fail.

```{r}
compadre_lambda <- Compadre %>%
  cdb_flag() %>%
  filter(check_NA_A == FALSE) %>% # remove matrices with missing values
  mutate(mat_A = matA(.)) %>% # extract list-column of matA
  mutate(lam = sapply(mat_A, popdemo::eigs, what = "lambda"))
```

In the example above, `sapply()` returns a scalar value for every row of the database. If we instead want to derive a more complex object for every row such as a vector or matrix, we can use the function `lapply()` which always returns a list. In the example below we use `lapply()` to calculate vectors of stage-specific survival (column sums of `matU`) for every row of the database.

```{r}
compadre_stage_surv <- Compadre %>%
  mutate(stage_survival = lapply(matU(.), colSums))

# print vector of stage-specific survival for 20th row
compadre_stage_surv$stage_survival[[20]]
```

#### Vectorizing over multiple arguments

Let's say we want to know the survival probability for the first 'active' stage class (i.e. a stage that's not dormant or propagule) for every row of the database. One approach is to write our own function that takes `matU` and the integer index of the first active stage class, and returns the corresponding survival probability, e.g.

```{r}
SurvFirstActive <- function(matU, first_active) colSums(matU)[first_active]
```

Notice that this function has two arguments, both of which will vary across rows of the database. We therefore need to vectorize this function over both arguments, which we can do with `mapply()` (i.e. "multivariate apply"). Here's an example:

```{r}
compadre_surv_first_active <- Compadre %>%
  mutate(
    surv_1 = mapply(
      FUN = SurvFirstActive, # function
      matU = matU(.), # argument 1
      first_active = mpm_first_active(.)
    ) # argument 2
  )
```


## The `group_by` and `summarize` functions

The `group_by()` and `summarize()` functions are used for split-apply-combine operations, where we want to apply a function separately to different groups within our data, and then recombine the results.

Here's an example. We'll use `group_by()` to group our database by unique values of species (column `SpeciesAccepted`), and then use `summarize()` to count the number of unique populations (column `MatrixPopulation`) for each group. Notice that the output is a regular tibble (i.e. no longer a `CompadreDB` object) with one row for each unique 'group' (i.e. species).

```{r}
# count number of unique populations by species
Compadre %>%
  group_by(SpeciesAccepted) %>%
  summarize(n_populations = length(unique(MatrixPopulation))) %>%
  arrange(desc(n_populations)) # arrange in descending order of n_pops
```

If, instead of a summary table with one row per species, we would rather append the number of populations per species as a new column of our `CompadreDB` object, then we can follow `group_by()` with `mutate()` instead of `summarize()`. Here's an example:

```{r}
# subset to species with 10+ unique populations
compadre_replicated_pops <- Compadre %>%
  group_by(SpeciesAccepted) %>%
  mutate(n_pops = length(unique(MatrixPopulation))) %>%
  ungroup() %>%
  subset(n_pops >= 10)
```

Make sure to use `ungroup()` after you're finished with the groups, to prevent unexpected behaviour later on.

## Obtaining a single representative of each species

It is a common task in comparative analyses to obtain a single representative of each species. One way that users can do that is by using `group_by()`, followed by `slice()` and `sample()` to randomly select a representative from each accepted species. 

```{r}
singleRepresentativeSpecies <- Compadre %>%
  group_by(SpeciesAccepted) %>%
  slice(sample(1))
```


## Using CompadreDB objects with `ggplot`

Rcompadre includes methods that allow CompadreDB objects to be used as data arguments within `ggplot()`, just like any old data frame. Below are some example plots.

__Plot study coordinates on world map:__

```{r warning=FALSE, fig.width = 6, fig.height = 4}
#| fig.alt: >
#|   A world map showing geographic distribution of COMPADRE data. 
ggplot2::ggplot(Compadre, aes(Lon, Lat)) +
  borders(database = "world", fill = "grey80", col = NA) +
  geom_point(col = "steelblue", size = 1.8, alpha = 0.8)
```

__Boxplots of life expectancy from the first non-propagule stage, arranged by OrganismType, based on 'grand mean matrices' from studies of wild, unmanipulated populations:__

This one is rather advanced, combining all of the topics in this vignette. So don't worry if you can't immediately understand every line.

```{r warning=FALSE, fig.width = 6, fig.height = 4}
#| fig.alt: >
#|   Boxplots of life expectancy from the first non-propagule stage, 
#|   arranged by OrganismType, based on 'grand mean matrices' from 
#|   studies of wild, unmanipulated populations 

# function to calculate life expectancy
lifeExpectancy <- function(matU, startLife) {
  N <- solve(diag(nrow(matU)) - matU)
  return(colSums(N)[startLife])
}

compadre_life_expect <- Compadre %>%
  filter(
    MatrixComposite != "Seasonal", # filter is the dplyr version of subset
    MatrixTreatment == "Unmanipulated",
    MatrixCaptivity == "W",
    ProjectionInterval == "1"
  ) %>%
  mutate(StageID = cdb_id_stages(.)) %>%
  cdb_collapse(columns = "StageID") %>%
  cdb_flag() %>%
  filter(
    check_NA_U == FALSE,
    check_zero_U == FALSE,
    check_singular_U == FALSE
  ) %>%
  mutate(matU = matU(.), start_life = mpm_first_active(.)) %>%
  mutate(life_expectancy = mapply(lifeExpectancy, matU, start_life)) %>%
  filter(life_expectancy >= 1) %>%
  mutate(OrganismType = reorder(OrganismType, life_expectancy, median))

ggplot2::ggplot(compadre_life_expect, aes(OrganismType, life_expectancy)) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip() +
  labs(x = NULL, y = "Life expectancy (years)")
```
