---
title: "Vectorising with Rcompadre"
author: "Patrick Barks"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vectorising with Rcompadre}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setupDarwin, include=FALSE, eval = Sys.info()[["sysname"]] == "Darwin"}
# The following line seems to be required by pkgdown::build_site() on my machine, but causes build to break with R-CMD-CHECK on GH
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

## Introduction

COM(P)ADRE databases contain thousands of matrix population models. If we want to derive traits from a large set of these matrices, we'll need to use either loops or vectorisation.

Vectorising means applying a function to each element of a vector. Here vector is defined broadly — it could be a sequence of character strings, a column of a `data.frame`, a `list` of matrices, etc. Vectorised code generally runs faster than loops, and many R users find that vectorised code is easier to write and understand.

## Preliminaries

We'll start by loading a few packages and a dataset that we'll be using throughout this vignette. The dataset `Compadre` is a subset of a recent COMPADRE release that's built into `Rcompadre`.

```{r}
library(Rcompadre)
library(popdemo)
data(Compadre)
```

## Introduction to vectorisation

To understand vectorisation, we first need a vector. For this purpose, we'll extract a list of __A__ matrices from the `mat` column of `Compadre`, and add this list of matrices to `Compadre` as a new column.

```{r}
Compadre$matA <- matA(Compadre)
```

This new column, `Compadre$matA`, is both a vector and a list.

```{r}
is.vector(Compadre$matA) # it really is a vector
is.list(Compadre$matA) # and also a list
length(Compadre$matA) # with 150 matrices
Compadre$matA[1:3] # here are the first three
```

Let's say we want to calculate the dimension of every matrix in `Compadre$matA`. In fact, `Compadre` already has this data in the column 'MatrixDimension', but let's say we want to double-check it. We'll use the function `nrow()`, and assume that the number of rows and columns are equal. But we can't use `nrow()` directly on `Compadre$matA`, because the function `nrow()` isn't vectorised. It can only take one object at a time.


#### Manual approach

We could do something like this...

```{r}
Compadre$dim <- numeric(nrow(Compadre)) # create empty vector to store output
Compadre$dim[1] <- nrow(Compadre$matA[[1]]) # nrow matrix 1
Compadre$dim[2] <- nrow(Compadre$matA[[2]]) # nrow matrix 2
Compadre$dim[3] <- nrow(Compadre$matA[[3]]) # nrow matrix 3
# ... all the way to 150
```

But that's not very efficient for 150 matrices.

#### Loop approach

A loop would be much more efficient here.

```{r}
# create empty vector to store output
Compadre$dim <- numeric(nrow(Compadre))

# loop through all rows of Compadre
for (i in seq_len(nrow(Compadre))) {
  Compadre$dim[i] <- nrow(Compadre$matA[[i]])
}
```

#### Vectorised approach

An even nicer approach is to _vectorise_ the function `nrow()` over the vector `Compadre$matA` using `sapply()`.

```{r}
Compadre$dim <- sapply(Compadre$matA, nrow)
```

`sapply()` applies the function specified in the 2nd argument (`nrow()`) to every element of the vector in the first argument (`Compadre$matA`), and returns a vector of the results. The advantage of `sapply()` over the loop is that we don't need to pre-define an object to store the results.

#### Vectorising custom functions

We can also vectorise with a custom function. Let's say we want to know, for every matrix, whether there are stages with no transitions (i.e. any column sums equal to zero). Here's a vectorised approach.

```{r}
# function to determine whether matrix 'mat' has any stages with no transitions
NullStages <- function(mat) any(colSums(mat) == 0)

# apply function to every element of A
Compadre$null_stages <- sapply(Compadre$matA, NullStages)
```

The key to vectorising is to make sure the function works on individual elements of the vector.

```{r, eval=FALSE}
NullStages(Compadre$matA[[1]]) # apply function to single element
```

## Accessor functions and vectorisation

Note that, in the example above, it wasn't necessary to create the column `Compadre$matA` before vectorising over the __A__ matrices. We could have simply used the `matA()` accessor within `sapply()`. 

```{r}
Compadre$null_stages <- sapply(matA(Compadre), NullStages)
```

#### Using `cdb_unnest()` to avoid accessors

That said, using accessor funtions can get tedious. Rather than constantly using accessor functions to extract components of the `mat` column, we could use the function `cdb_unnest()` to extract separate columns for all matrix components at the start of our analysis.

```{r}
# create new columns matA, matU, matF, matC, MatrixClassAuthor, etc..
CompUnnest <- cdb_unnest(Compadre)
```

Then we can refer to any component using `$`, e.g.

```{r}
# apply NullStages to every matA
CompUnnest$null_stages <- sapply(CompUnnest$matA, NullStages)

# count number of dormant stages in every MatrixClassOrganized
NumberDormant <- function(stages) length(which(stages == "dorm"))
CompUnnest$n_dormant <- sapply(CompUnnest$MatrixClassOrganized, NumberDormant)
```


## Other apply functions

`vapply()` is similar to `sapply()`, except that the output type is specified as an argument.

```{r}
sapply(CompUnnest$matA[1:6], nrow)
vapply(CompUnnest$matA[1:6], nrow, numeric(1)) # must specify output type
```

`lapply()` always returns a `list`, so it's useful if our output is more complex than a single value for each input. For example, we could use `lapply` to calculate vectors of stage-specific survival (column sums of `matU`).

```{r}
lapply(CompUnnest$matU[1:4], function(m) colSums(m))
```

`mapply()` is for vectorising over multiple arguments. For example, the `lifeExpectancy()` function below (taken from the package [Rage](https://github.com/jonesor/Rage)) calculates life expectancy given two arguments: a __U__ matrix, and an integer indicator for the stage class reflecting the 'start of life'. The start of life is often defined as the first 'active' stage class (i.e. not propagule or dormant), the index of which will vary from row to row.

```{r}
# function to calculate life expectancy
lifeExpectancy <- function(matU, startLife) {
  N <- solve(diag(nrow(matU)) - matU)
  return(colSums(N)[startLife])
}

# get index of first active stage class with mpm_first_active()
CompUnnest$start_life <- mpm_first_active(CompUnnest)

# vectorise lifeExpectancy over matU and start_life
mapply(
  lifeExpectancy, # function
  CompUnnest$matU[1:6], # first argument to vectorise over
  CompUnnest$start_life[1:6]
) # second argument to vectorise over
```


## When functions fail

Just like loops, vectorisation fails if the function being vectorised throws an error on _any_ element of the vector. Here's an example. The `eigs()` function from [popdemo](https://CRAN.R-project.org/package=popdemo) calculates the expected population growth rate given a projection matrix. It'll work on most of the __A__ matrices in `Compadre`, but fails on matrices that contain missing values (`NA`).

```{r error=TRUE}
# works for a single matrix
popdemo::eigs(CompUnnest$matA[[1]], what = "lambda")

# but fails when applied to all matrices because a few have missing values
CompUnnest$lambda <- sapply(CompUnnest$matA, popdemo::eigs, what = "lambda")
```

There are two basic approaches to overcoming this:

#### 1. Remove or skip problem elements

If `eigs()` doesn't work on matrices with missing values, one approach is to simply remove matrices with missing values. The `cdb_flag()` function is an easy way to check for missing values, and other common issues that might hinder our analyses.

```{r}
# add column 'check_NA_A', indicating whether matA contains missing values (T/F)
CompFlag <- cdb_flag(CompUnnest, checks = "check_NA_A")

# remove rows where matA contains missing values
CompSub <- subset(CompFlag, check_NA_A == FALSE)

# apply lambda() to every remaining matA
CompSub$lambda <- sapply(matA(CompSub), popdemo::eigs, what = "lambda")
```

Alternatively, if we want to avoid subsetting, we could pre-define a placeholder column for the result, and then selectively apply the `eigs()` function to only those matrices that don't contain missing values.

```{r}
# identify rows with no missing values in matA
no_missing <- which(CompFlag$check_NA_A == FALSE)

# create placeholder column for lambda
CompFlag$lambda <- NA

# apply eigs() to all matA with no missing values
CompFlag$lambda[no_missing] <- sapply(CompFlag$matA[no_missing], popdemo::eigs, what = "lambda")
```

Rows where there were missing values in `matA` retain the original placeholder value of `NA`.

#### 2. Modify the function

A second approach is to modify the function we want to vectorise with so that it can natively handle special cases. For example, we might modify `eigs()` so that it returns `NA` if a matrix contains missing values.

```{r}
lambdaFn1 <- function(mat) {
  # check mat for missing values: if TRUE return NA, else return eigs(mat)
  ifelse(any(is.na(mat)), NA, popdemo::eigs(mat, what = "lambda"))
}

CompUnnest$lambda <- sapply(CompUnnest$matA, lambdaFn1)
```

If the special cases are harder to test for, we could use R's condition-handling functions like `try()` or `tryCatch()`. Here's an example.

```{r}
lambdaFn2 <- function(mat) {
  # try eigs(mat): if error return NA
  tryCatch(eigs(mat, what = "lambda"), error = function(err) NA)
}

CompUnnest$lambda <- sapply(CompUnnest$matA, lambdaFn2)
```

This latter approach requires caution, as we'll get an `NA` for _any_ error. Some errors might reflect problems with our data or code that are fixable, in which case an `NA` may be misleading.
