---
title: "Age-from-stage analyses"
author: "Patrick Barks"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age-from-stage analyses}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Regardless of whether the stage classes of a matrix population model (MPM) are based on age, size, and/or ontogeny, it's possible to obtain age-specific schedules of survivorship (lx) and reproduction (mx) using 'age-from-stage' methods, as described by Caswell (2001).

## Preliminaries

We'll start by loading a few packages and a dataset that we'll be using throughout this vignette. The dataset `Compadre` is a subset of a recent COMPADRE release that's built into Rcompadre.

```{r}
library(Rcompadre)
library(Rage)
data(Compadre)
options(tibble.width = Inf) # always print all columns of tibble
```

## Age-from-stage methods with a single matrix population model

Next we'll grab an MPM, arbitrarily selecting the fourth one in `Compadre`.

```{r}
mpm <- Compadre$mat[[4]]
mpm
```

This MPM has 5 stage class, and it's apparent from the `MatrixClassAuthor` column that the stages are not based solely on age. Nonetheless, we can estimate age-schedules of survivorship and reproduction using the functions `ageSpecificSurv()` and `ageSpecificRepro()` from the library [Rage](https://github.com/jonesor/Rage).

```{r}
# extract U and F matrices
mat_U <- matU(mpm)
mat_F <- matF(mpm)

# calculate lx
lx <- ageSpecificSurv(mat_U, startLife = 1, N = 15)

# calculate mx
mx <- ageSpecificRepro(mat_U, mat_F, startLife = 1, N = 15)
```

In addition to the relevant matrix components, the `ageSpecific-` functions require two extra arguments. The first, `startLife`, is an integer indicating which stage reflects the 'start of life'. Usually this will be `1`, but sometimes we might want to skip over stages that are propagule (i.e. seed) or dormant. The MPM we selected didn't have any such stages (see the column `MatrixClassOrganized` above), so we can stick with `1`. The second argument, `N` is the number of time steps to calculate over.

Let's take a look at the trajectories.

```{r, fig.width = 6, fig.height = 4}
plot(lx, ylim = c(0, 1), type = "l", xlab = "Age")
plot(mx, type = "l", xlab = "Age")
```

## Extending to many matrix population models

Now we'll extend the basic approach above to many models. Specifically, we'll examine trajectories of survivorship for all of the tree species in `Compadre`.

First, we'll subset `Compadre` to our group of interest (`OrganismType == "Tree"`). We'll also remove matrices with missing values, and limit our selection to matrices with a periodicity (i.e. transition interval) of 1 year.

```{r}
comp_flag <- cdb_flag(Compadre, "check_NA_U")

comp_use <- subset(comp_flag, OrganismType == "Tree" &
                     check_NA_U == FALSE &
                     AnnualPeriodicity == 1)
```

Let's take a look at the species/populations that made the cut.

```{r}
CompadreData(comp_use)[,c("SpeciesAccepted", "MatrixPopulation", "MatrixTreatment")]
```

Notice that there are 3 matrices for the species _Phyllanthus indofischeri_, reflecting different treatment groups. Let's collapse these replicates down to a single matrix per species, by averaging the relevant MPMs using `cdb_collapse()`. We'll also use the function `cdb_id_stages()`, to make sure we're only collapsing matrices that have the same stage class definitions.

```{r}
# add column ID-ing matrices with same MatrixClassAuthor vector
comp_use$stage_id <- cdb_id_stages(comp_use)

# collapse database to single matrix per species * MatrixClassAuthor
comp_collapse <- cdb_collapse(comp_use, "stage_id")

# check species/populations again
CompadreData(comp_collapse)[,c("SpeciesAccepted", "MatrixPopulation", "MatrixTreatment")]
```

Next, let's look at the organized stage classes for each MPM. If any of our MPMs include propagule or dormant stage classes, we may want to account for them when calculating lx.

```{r}
MatrixClassOrganized(comp_collapse)
```

Indeed, 1 MPM incudes a propagule stage. So let's use the function `mpm_first_active()` to determine the first 'active' stage class for each MPM, which we'll use to define the start of life.

```{r}
comp_collapse$start_life <- mpm_first_active(comp_collapse)
```

Finally, we'll use `mapply()` to vectorize the function `ageSpecificSurv()` over arguments matU and startLife. By default, `mapply()` will return an array where the columns correspond to rows of our database, and rows correspond to ages — 0 through `N`. Given that the output is an array, we'll use the function `matplot()` to plot age-trajectories of survivorship for each species.

```{r, fig.width = 6, fig.height = 4}
lx_array <- mapply(ageSpecificSurv,
                   matU = matU(comp_collapse),
                   startLife = comp_collapse$start_life,
                   N = 250)

matplot(lx_array, type = "l", lty = 1, log = "y", ylim = c(0.0001, 1),
        lwd = 1.5, xlab = "Age (years)", ylab = "lx")
```

## References

Caswell, H. (2001). Matrix Population Models: Construction, Analysis, and Interpretation. 2nd edition. Sinauer Associates, Sunderland, MA. ISBN-10: 0878930965
